Location and Scale-Invariant Power Transformations for Transforming Data to Normality

### 2025-04-11 manuscript_latex/manuscript.tex
Updated tex file of main manuscript

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
%% \usepackage{amsmath,amssymb,amsfonts}%
%% \usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
%% \usepackage{algorithm}%
%% \usepackage{algorithmicx}%
%% \usepackage{algpseudocode}%
\usepackage{listings}%
%%%%

\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator{\sgn}{sgn}

@@

\affil*[1]{
	\orgdiv{National Center for Tumor Diseases Dresden (NCT/UCC)},
	\orgname{German Cancer Research Center (DKFZ), Faculty of Medicine and University Hospital Carl Gustav Carus, TUD Dresden University of Technology, Helmholtz-Zentrum Dresden-Rossendorf (HZDR)},
	\orgaddress{\street{Fetscherstra{\ss}e 74/PF 64}, \city{Dresden}, \postcode{01307}, \country{Germany}}
}

@@

\abstract{Power transformations are used to stabilize variance and achieve
normality of features, especially in methods assuming normal 
distributions such as ANOVA and linear discriminant analysis. 
However, the commonly used Box-Cox and Yeo-Johnson power transformation 
methods are sensitive to the location, scale, and presence of 
outliers in the data. Here we present location- and scale-invariant Box-Cox and Yeo-Johnson
transformations to mitigate these issues. We derive maximum likelihood
estimation criteria for optimizing transformation parameters and propose
robust adaptations that reduce the influence of outliers. 
We also introduce an empirical test for assessing central normality
of transformed features. In simulations and real-world datasets, robust location- and scale-invariant 
transformations outperform conventional variants, resulting in better 
transformations to central normality. In a machine learning experiment
with 232 datasets with numerical features, integrating robust 
location- and scale-invariant power transformations into an automated data 
processing and machine learning pipeline did not result in a meaningful
improvement or detriment to model performance compared to conventional variants.
In conclusion, robust location- and scale-invariant power
transformations can replace conventional variants.}

@@

Many statistical and some machine learning methods assume normality of
the underlying data, e.g.~analysis of variance and linear discriminant
analysis. However, numerical features in datasets may strongly deviate
from normal distributions, e.g.~by being skewed. Power transformations
aim to stabilise variance and improve normality of such features
\citep{Bartlett1947-rx, Tukey1957-rt}. The two most commonly used
transformations are that of \citet{Box1964-mz} and \citet{Yeo2000-vw}.
The Box-Cox transformation of a feature value \(x_i\) of feature
\(\mathbf{X}=\left\{x_1, x_2, \ldots, x_n \right\}\) under the
transformation parameter \(\lambda\) is defined as:

\begin{equation}
\label{eqn:box-cox-original}
\phi_{\text{BC}}^\lambda (x_i) = 
\begin{cases}
\left(x_i^\lambda - 1 \right) / \lambda & \text{if } \lambda \neq 0\\
\log(x_i) & \text{if } \lambda = 0
\end{cases}
\end{equation}

One limitation of the Box-Cox transformation is that it is only defined
for \(x_i > 0\). In contrast, the Yeo-Johnson transformation under the
transformation parameter \(\lambda\) is defined for any
\(x_i \in \mathbb{R}\):

\begin{equation}
\label{eqn:yeo-johnson-original}
\phi_{\text{YJ}}^\lambda (x_i) = 
\begin{cases}
\left( \left( 1 + x_i \right)^\lambda - 1\right) / \lambda & \text{if } \lambda \neq 0 \text{ and } x_i \geq 0\\
\log(1 + x_i) & \text{if } \lambda = 0 \text{ and } x_i \geq 0\\
-\left( \left( 1 - x_i\right)^{2 - \lambda} - 1 \right) / \left(2 - \lambda \right) & \text{if } \lambda \neq 2 \text{ and } x_i < 0\\
-\log(1 - x_i) & \text{if } \lambda = 2 \text{ and } x_i < 0
\end{cases}
\end{equation}

The \(\lambda\)-parameter is typically optimised using maximum
likelihood estimation under the assumption that the transformed feature
is normally distributed. As noted by Raymaekers and Rousseeuw, this
approach is sensitive to outliers, and robust versions of Box-Cox and
Yeo-Johnson transformations were devised \citep{Raymaekers2024-zf}.

@@

Statistical tests for normality, such as the Shapiro-Wilk test
\citep{Shapiro1965-zd}, could be automatically applied to transformed features.
However, given sufficiently large sample sizes such tests can detect
trivial deviations from normality, and may lead to rejection of
sufficiently good power transformations.

@@

Box-Cox and Yeo-Johnson transformations are modified by introducing
shift parameter \(x_0\) and scale parameter \(s\) into equations
\ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}. The
location- and scale-invariant Box-Cox transformation of a feature value
\(x_i\) of feature \(\mathbf{X}\) under transformation parameter
\(\lambda\), shift parameter \(x_0\) and scale parameter \(s\) is then:

\begin{equation}
\label{eqn:box-cox-invariant}
\phi_{\text{BC}}^{\lambda, x_0, s} (x_i) = 
\begin{cases}
\left( \left(\frac{x_i - x_0}{s} \right)^\lambda - 1 \right) / \lambda & \text{if } \lambda \neq 0\\
\log\left[\frac{x_i - x_0}{s}\right] & \text{if } \lambda = 0
\end{cases}
\end{equation}

where \(x_i - x_0 > 0\). Likewise, the location- and scale-invariant
Yeo-Johnson transformation of a feature value \(x_i\) under
transformation parameter \(\lambda\), shift parameter \(x_0\) and scale
parameter \(s\) is:

\begin{equation}
\label{eqn:yeo-johnson-invariant}
\phi_{\text{YJ}}^{\lambda, x_0, s} (x_i) = 
\begin{cases}
\left( \left( 1 + \frac{x_i - x_0}{s}\right)^\lambda - 1\right) / \lambda & \text{if } \lambda \neq 0 \text{ and } x_i - x_0 \geq 0\\
\log\left[1 + \frac{x_i - x_0}{s}\right] & \text{if } \lambda = 0 \text{ and } x_i - x_0 \geq 0\\
-\left( \left( 1 - \frac{x_i - x_0}{s}\right)^{2 - \lambda} - 1 \right) / \left(2 - \lambda \right) & \text{if } \lambda \neq 2 \text{ and } x_i - x_0 < 0\\
-\log\left[1 - \frac{x_i - x_0}{s}\right] & \text{if } \lambda = 2 \text{ and } x_i - x_0 < 0
\end{cases}
\end{equation}

For both invariant transformations, \(\lambda\), \(x_0\) and \(s\)
parameters can be obtained by maximising the log-likelihood function,
i.e.~using maximum likelihood estimation (MLE). A full derivation of the
log-likelihood function for both transformations is shown in \ref{appendix-a-log-likelihood-functions-for-location-and-scale-invariant-power-transformation}.
The location- and scale-invariant Box-Cox log-likelihood function is:

\begin{equation}
\label{eqn:box-cox-invariant-log-likelihood}
\begin{split}
\mathcal{L}_{\text{BC}}^{\lambda, x_0, s} = & -\frac{n}{2} \log \left[2 \pi \sigma^2 \right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi_{BC}^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& -n \lambda \log s + \left( \lambda - 1 \right) \sum_{i=1}^n \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}

subject to \(x_i - x_0 > 0\). \(\mu\) and \(\sigma^2\) are the mean and
variance of the Box-Cox transformed feature
\(\phi_{\text{BC}}^{\lambda, x_0, s} (\mathbf{X})\), respectively.
Similarly, the location- and scale-invariant Yeo-Johnson log-likelihood
function is:

\begin{equation}
\label{eqn:yeo-johnson-invariant-log-likelihood}
\begin{split}
\mathcal{L}_{\text{YJ}}^{\lambda, x_0, s} = & -\frac{n}{2} \log\left[2 \pi \sigma^2\right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi_{YJ}^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& - n \log s + (\lambda - 1) \sum_{i=1}^n \sgn(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}

where \(\mu\) and \(\sigma^2\) are the mean and variance of the
Yeo-Johnson transformed feature
\(\phi_{\text{YJ}}^{\lambda, x_0, s} (\mathbf{X})\), respectively.

@@

Real-world data may contain outliers, to which maximum likelihood
estimation can be sensitive. Their presence may lead to poor
transformations to normality, as shown in Figure
\ref{fig:decreased-normality}. As indicated by \citet{Raymaekers2024-zf},
the general aim of power transformations should be to transform
non-outlier data to normality, i.e.~achieve \emph{central normality}. To
achieve this, they devised an iterative procedure to find a robust
estimate of the transformation parameter \(\lambda\). Briefly, this
process requires identifying outliers in the data and weighting such
instances during the optimisation process. \citet{Raymaekers2024-zf}
achieve this through weighted maximum likelihood estimation.
However, because this procedure iteratively estimates and updates
\(\lambda\), it can not be used here to simultaneously estimate
\(\lambda\), \(x_0\) and \(s\) for location- and scale-invariant power
transformations. Nonetheless, as a procedure, weighted MLE can be used
for estimating the transformation, shift and scale parameters.

Here, weighted maximum likelihood estimation is based on equations
\ref{eqn:box-cox-invariant-log-likelihood} and
\ref{eqn:yeo-johnson-invariant-log-likelihood}. Compared to \citet{Raymaekers2024-zf},
these log-likelihood functions includes additional
terms to accommodate estimation of \(x_0\) and \(s\). The weighted
location- and scale-invariant Box-Cox log-likelihood function is:

\begin{equation}
\label{eqn:box-cox-weighted-invariant-log-likelihood}
\begin{split}
\mathcal{L}_{\text{rBC}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi_{\text{BC}}^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \lambda \left( \sum_{i=1}^n w_i \right) \log s + \left( \lambda - 1 \right) \sum_{i=1}^n w_i \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}

where \(\mu_w\) and \(\sigma^2_w\) are the weighted mean and weighted
variance of the Box-Cox transformed feature
\(\phi_{\text{BC}}^{\lambda, x_0, s} (\mathbf{X})\):

\begin{equation}
\sigma_w^2 = \frac{\sum_{i=1}^n w_i \left(\phi_{\text{BC}}^{\lambda, x_0, s} (x_i) - \mu_w \right)^2}{\sum_{i=1}^n w_i} \quad \text{with } \mu_w = \frac{\sum_{i=1}^n w_i \phi_{\text{BC}}^{\lambda, x_0, s} (x_i)} {\sum_{i=1}^n w_i}
\end{equation}

@@

\begin{equation}
\label{eqn:yeo-johnson-weighted-invariant-log-likelihood}
\begin{split}
\mathcal{L}_{\text{rYJ}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi_{\text{YJ}}^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \left( \sum_{i=1}^n w_i \right) \log s + (\lambda - 1) \sum_{i=1}^n w_i \sgn(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}

\begin{equation}
\sigma_w^2 = \frac{\sum_{i=1}^n w_i \left(\phi_{\text{YJ}}^{\lambda, x_0, s} (x_i) - \mu_w \right)^2}{\sum_{i=1}^n w_i} \quad \text{with } \mu_w = \frac{\sum_{i=1}^n w_i \phi_{\text{YJ}}^{\lambda, x_0, s} (x_i)} {\sum_{i=1}^n w_i}
\end{equation}

@@

\begin{itemize}
\item
  A step function, with \(\delta_1 \geq 0\) as threshold parameter:
\end{itemize}

\begin{equation}
w_i =
\begin{cases}
1 & \text{if } \left| \dot{x}_i \right| \leq \delta_1\\
0 & \text{if } \left| \dot{x}_i \right| > \delta_1
\end{cases}
\end{equation}

\begin{itemize}
\item
  A triangle function (or generalised Huber weight), with
  \(\delta_1 \geq 0\) and \(\delta_2 \geq \delta_1\) as threshold
  parameters:
\end{itemize}

\begin{equation}
w_i =
\begin{cases}
1 & \text{if } \left| \dot{x}_i \right| < \delta_1\\
1 - \frac{\left| \dot{x}_i \right| - \delta_1}{\delta_2 - \delta_1} & \text{if } \delta_1 \leq \left| \dot{x}_i \right| \leq \delta_2 \\
0 & \text{if } \left| \dot{x}_i \right| > \delta_2
\end{cases}
\end{equation}

\begin{itemize}
\item
  A tapered cosine function \citep{Tukey1967-eb}, with \(\delta_1 \geq 0\)
  and \(\delta_2 \geq \delta_1\) as threshold parameters:
\end{itemize}

\begin{equation}
w_i =
\begin{cases}
1 & \text{if } \left| \dot{x}_i \right| < \delta_1\\
0.5 + 0.5 \cos\left(\pi \frac{\left| \dot{x}_i \right| - \delta_1}{\delta_2 - \delta_1} \right) & \text{if } \delta_1 \leq \left| \dot{x}_i \right| \leq \delta_2 \\
0 & \text{if } \left| \dot{x}_i \right| > \delta_2
\end{cases}
\end{equation}

@@

\caption{Weighting functions investigated in this study to make power transformations more robust against outliers. In this example, the step function was parametrised with $\delta_1 = 0.60$. The triangle and tapered cosine functions were both parametrised with $\delta_1 = 0.30$ and $\delta_2 = 0.90$.}\label{fig:weighting-functions}
\end{figure}

@@

\begin{itemize}
\item
  The weighting function uses empirical probabilities of the
  distribution of the original feature \(\mathbf{X}\). After sorting
  \(\mathbf{X}\) in ascending order, probabilities are determined as
  \(p_i = \frac{i - 1/3}{n + 1/3}\), with \(i = 1, 2, \ldots n\), with
  \(n\) the number of instances of feature \(\mathbf{X}\). Then
  \(\dot{x}_i = p^{*}_i=2 \left( p_i - 0.5\right)\), so that argument is
  zero-centered.
\item
  The weighting function uses the z-score of the transformed feature
  \(\phi^{\lambda, x_0, s} (\mathbf{X})\). After \citep{Raymaekers2024-zf},
  \(z_i = \frac{\phi^{\lambda, x_0, s}(x_i) - \mu_M}{\sigma_M}\). Here,
  \(\mu_M\) and \(\sigma_M\) are robust Huber M-estimates of location
  and scale of the transformed feature
  \(\phi^{\lambda, x_0, s} (\mathbf{X})\) \citep{Huber1981-su}. Then
  \(\dot{x}_i = z_i\).
\item
  After sorting \(\mathbf{X}\) in ascending order, the weighting
  function uses the residual error between the z-score of the
  transformed feature \(\phi^{\lambda, x_0, s} (\mathbf{X})\) and the
  theoretical z-score from a standard normal distribution:
  \(r_i =\left| \left( \phi^{\lambda, x_0, s}(x_i) - \mu_M)\right) / \sigma_M - F^{-1}_{\mathcal{N}}(p_i) \right|\),
  with \(\mu_M\), \(\sigma_M\) and \(p_i\) as defined above. Then
  \(\dot{x}_i = r_i\).
\end{itemize}

>>>>

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
%% \usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
%% \usepackage{algorithm}%
%% \usepackage{algorithmicx}%
%% \usepackage{algpseudocode}%
\usepackage{listings}%
%%%%
\newtheorem*{definition}{Definition}
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator{\sgn}{sgn}

@@

\affil*[1]{
	\orgdiv{National Center for Tumor Diseases (NCT), NCT/UCC Dresden },
	\orgname{a partnership between DKFZ, Faculty of
	Medicine and University Hospital Carl Gustav Carus, TUD Dresden University of Technology, and
	Helmholtz-Zentrum Dresden-Rossendorf (HZDR)},
	\orgaddress{\street{Fetscherstra{\ss}e 74/PF 64}, \city{Dresden}, \postcode{01307}, \country{Germany}}
}

@@

\abstract{Power transformations are used to stabilize variance and achieve
normality of features, especially in methods assuming normal
distributions such as ANOVA and linear discriminant analysis. However,
the commonly used Box-Cox and Yeo-Johnson power transformation methods
are sensitive to the location, scale, and presence of outliers in the
data. Here we present location- and scale-invariant Box-Cox and
Yeo-Johnson transformations to mitigate these issues. We derive maximum
likelihood estimation criteria for optimizing transformation parameters
and propose robust adaptations that reduce the influence of outliers. We
also introduce an empirical test for assessing central normality of
transformed features. In simulations and real-world datasets, robust
location- and scale-invariant transformations outperform conventional
variants, resulting in better transformations to central normality. In a
machine learning experiment with 231 datasets with numerical features,
integrating robust location- and scale-invariant power transformations
into an automated data processing and machine learning pipeline did not
result in a meaningful improvement or detriment to model performance
compared to conventional variants. In conclusion, robust location- and
scale-invariant power transformations can replace conventional variants.}

@@

Many statistical and some machine learning methods assume normality of
the underlying data, e.g.~analysis of variance and linear discriminant
analysis. However, numerical features in datasets may strongly deviate
from normal distributions, e.g.~by being skewed. Power transformations
aim to stabilise variance and improve normality of such features
(Bartlett 1947; John W. Tukey 1957). The two most commonly used
transformations are that of Box and Cox (1964) and Yeo and Johnson
(2000).

The Box-Cox transformation is defined as:

\begin{definition}[Box-Cox power transformation]
Let $\mathbf{X} = \left\{x_1, x_2, \ldots, x_n | x_i > 0 \right\}$ be a finite sequence of length $n > 0$.
Let $\lambda \in \mathcal{R}$ be a transformation parameter.
Then the Box-Cox power transformation of element $x_i$ is defined as:
\begin{equation}
\label{eqn:box-cox-original}
\phi_{\text{BC}}^\lambda (x_i) = 
\begin{cases}
\left(x_i^\lambda - 1 \right) / \lambda & \text{if } \lambda \neq 0\\
\log(x_i) & \text{if } \lambda = 0
\end{cases}
\end{equation}
\end{definition}

One limitation of the Box-Cox transformation is that it is only defined
for \(x_i > 0\). In contrast, the Yeo-Johnson transformation under the
transformation parameter \(\lambda\) is defined for any
\(x_i \in \mathbb{R}\), as follows:

\begin{definition}[Yeo-Johnson power transformation]
Let $\mathbf{X} = \left\{x_1, x_2, \ldots, x_n | x_i \in \mathbb{R} \right\}$ be a finite sequence of length $n > 0$.
Let $\lambda \in \mathcal{R}$ be a transformation parameter.
\begin{equation}
\label{eqn:yeo-johnson-original}
\phi_{\text{YJ}}^\lambda (x_i) = 
\begin{cases}
\left( \left( 1 + x_i \right)^\lambda - 1\right) / \lambda & \text{if } \lambda \neq 0 \text{ and } x_i \geq 0\\
\log(1 + x_i) & \text{if } \lambda = 0 \text{ and } x_i \geq 0\\
-\left( \left( 1 - x_i\right)^{2 - \lambda} - 1 \right) / \left(2 - \lambda \right) & \text{if } \lambda \neq 2 \text{ and } x_i < 0\\
-\log(1 - x_i) & \text{if } \lambda = 2 \text{ and } x_i < 0
\end{cases}
\end{equation}
\end{definition}

The \(\lambda\)-parameter is typically optimised using maximum
likelihood estimation under the assumption that the transformed feature
is normally distributed. As noted by Raymaekers and Rousseeuw, this
approach is sensitive to outliers, and robust versions of Box-Cox and
Yeo-Johnson transformations were devised (Raymaekers and Rousseeuw
2024).

@@

Statistical tests for normality, such as the Shapiro-Wilk test (Shapiro
and Wilk 1965), could be automatically applied to transformed features.
However, given sufficiently large sample sizes such tests can detect
trivial deviations from normality, and may lead to rejection of
sufficiently good power transformations.

@@

Box-Cox and Yeo-Johnson transformations are modified by introducing
shift parameter \(x_0\) and scale parameter \(s\) into equations
\ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}. The
location- and scale-invariant Box-Cox transformation of a feature value
\(x_i\) of feature \(\mathbf{X}\) under transformation parameter
\(\lambda\), shift parameter \(x_0\) and scale parameter \(s\) is then
defined as:

\begin{definition}[Location- and scale- invariant Box-Cox power transformation]
Let $\lambda \in \mathcal{R}$ be a transformation parameter.
Let $x_0 \in \mathcal{R}$ and $s > 0$ be location and scale parameters.
Let $\mathbf{X} = \left\{x_1, x_2, \ldots, x_n | x_i - x_0 > 0 \right\}$ be a finite sequence of length $n > 0$.

Then the location- and scale-invariant Box-Cox power transformation of element $x_i$ is defined as:

\begin{equation}
\label{eqn:box-cox-invariant}
\phi_{\text{BC}}^{\lambda, x_0, s} (x_i) = 
\begin{cases}
\left( \left(\frac{x_i - x_0}{s} \right)^\lambda - 1 \right) / \lambda & \text{if } \lambda \neq 0\\
\log\left[\frac{x_i - x_0}{s}\right] & \text{if } \lambda = 0
\end{cases}
\end{equation}
\end{definition}

Likewise, the location- and scale-invariant Yeo-Johnson transformation
of a feature value \(x_i\) under transformation parameter \(\lambda\),
shift parameter \(x_0\) and scale parameter \(s\) is defined as:

\begin{definition}[Location- and scale- invariant Yeo-Johnson power transformation]
Let $\lambda \in \mathcal{R}$ be a transformation parameter.
Let $x_0 \in \mathcal{R}$ and $s > 0$ be location and scale parameters.
Let $\mathbf{X} = \left\{x_1, x_2, \ldots, x_n | x_i \in \mathcal{R} \right\}$ be a finite sequence of length $n > 0$.

Then the location- and scale-invariant Yeo-Johnson power transformation of element $x_i$ is defined as:

\begin{equation}
\label{eqn:yeo-johnson-invariant}
\phi_{\text{YJ}}^{\lambda, x_0, s} (x_i) = 
\begin{cases}
\left( \left( 1 + \frac{x_i - x_0}{s}\right)^\lambda - 1\right) / \lambda & \text{if } \lambda \neq 0 \text{ and } x_i - x_0 \geq 0\\
\log\left[1 + \frac{x_i - x_0}{s}\right] & \text{if } \lambda = 0 \text{ and } x_i - x_0 \geq 0\\
-\left( \left( 1 - \frac{x_i - x_0}{s}\right)^{2 - \lambda} - 1 \right) / \left(2 - \lambda \right) & \text{if } \lambda \neq 2 \text{ and } x_i - x_0 < 0\\
-\log\left[1 - \frac{x_i - x_0}{s}\right] & \text{if } \lambda = 2 \text{ and } x_i - x_0 < 0
\end{cases}
\end{equation}
\end{definition}

For both invariant transformations, \(\lambda\), \(x_0\) and \(s\)
parameters can be obtained by maximising the log-likelihood function,
i.e.~using maximum likelihood estimation (MLE). A full derivation of the
log-likelihood function for both transformations is shown in Appendix A.
The location- and scale-invariant Box-Cox log-likelihood function is:

\begin{definition} [Log-likelihood function of the location- and scale- invariant Box-Cox power transformation]
Let $\mathbf{X}$, $\lambda$, $x_0$, and $s$ be defined as earlier (Eqn. \ref{eqn:box-cox-invariant}).
Let $\mu$ and $\sigma^2$ be the mean and variance of the transformed sequence $\phi_{\text{BC}}^{\lambda, x_0, s} (\mathbf{X})$, respectively.

Then the log-likelihood function of the location- and scale- invariant Box-Cox power transformation is defined as:
\begin{equation}
\label{eqn:box-cox-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{BC}}^{\lambda, x_0, s} = & -\frac{n}{2} \log \left[2 \pi \sigma^2 \right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi_{BC}^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& -n \lambda \log s + \left( \lambda - 1 \right) \sum_{i=1}^n \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
\end{definition}

Similarly, the location- and scale-invariant Yeo-Johnson log-likelihood
function is:

\begin{definition}[Log-likelihood function of the location- and scale- invariant Yeo-Johnson power transformation]
Let $\mathbf{X}$, $\lambda$, $x_0$, and $s$ be defined as earlier (Eqn. \ref{eqn:yeo-johnson-invariant}).
Let $\mu$ and $\sigma^2$ be the mean and variance of the transformed sequence $\phi_{\text{YJ}}^{\lambda, x_0, s} (\mathbf{X})$, respectively.

Then the log-likelihood function of the location- and scale- invariant Yeo-Johnson power transformation is defined as:
\begin{equation}
\label{eqn:yeo-johnson-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{YJ}}^{\lambda, x_0, s} = & -\frac{n}{2} \log\left[2 \pi \sigma^2\right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi_{YJ}^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& - n \log s + (\lambda - 1) \sum_{i=1}^n \mathop{\mathrm{sgn}}(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}
\end{definition}

@@

Real-world data may contain outliers, to which maximum likelihood
estimation can be sensitive. Their presence may lead to poor
transformations to normality, as shown in Figure
\ref{fig:decreased-normality}. As indicated by Raymaekers and Rousseeuw
(2024), the general aim of power transformations should be to transform
non-outlier data to normality, i.e.~achieve \emph{central normality}. To
achieve this, they devised an iterative procedure to find a robust
estimate of the transformation parameter \(\lambda\). Briefly, this
process requires identifying outliers in the data and weighting such
instances during the optimisation process. Raymaekers and Rousseeuw
(2024) achieve this through weighted maximum likelihood estimation.
However, because this procedure iteratively estimates and updates
\(\lambda\), it can not be used here to simultaneously estimate
\(\lambda\), \(x_0\) and \(s\) for location- and scale-invariant power
transformations. Nonetheless, as a procedure, weighted MLE can be used
for estimating the transformation, shift and scale parameters.

Here, weighted maximum likelihood estimation is based on equations
\ref{eqn:box-cox-invariant-log-likelihood} and
\ref{eqn:yeo-johnson-invariant-log-likelihood}. Compared to Raymaekers
and Rousseeuw (2024), these log-likelihood functions include additional
terms to accommodate estimation of \(x_0\) and \(s\). The weighted
location- and scale-invariant Box-Cox log-likelihood function is:

\begin{definition}[Weighted log-likelihood function of the location- and scale- invariant Box-Cox power transformation]
Let $\mathbf{X}$, $\lambda$, $x_0$, and $s$ be defined as earlier (Eqn. \ref{eqn:box-cox-invariant}).
Let $w_i \geq 0$ be the weight corresponding to each element of $\mathbf{X}$.
Let $\mu_w$ be the weighted mean of the Box-Cox transformed sequence:
\begin{equation*}
\mu_w = \frac{\sum_{i=1}^n w_i \phi_{\text{BC}}^{\lambda, x_0, s} (x_i)} {\sum_{i=1}^n w_i}
\end{equation*}

Let $\sigma^2_w$ be the weighted variance of the Box-Cox transformed sequence:
\begin{equation*}
\sigma_w^2 = \frac{\sum_{i=1}^n w_i \left(\phi_{\text{BC}}^{\lambda, x_0, s} (x_i) - \mu_w \right)^2}{\sum_{i=1}^n w_i}
\end{equation*}

Then, the weighted log-likelihood function of the location- and scale- invariant Box-Cox power transformation is:

\begin{equation}
\label{eqn:box-cox-weighted-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{rBC}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi_{\text{BC}}^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \lambda \left( \sum_{i=1}^n w_i \right) \log s + \left( \lambda - 1 \right) \sum_{i=1}^n w_i \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
\end{definition}

@@

\begin{definition}[Weighted log-likelihood function of the location- and scale- invariant Yeo-Johnson power transformation]

Let $\mathbf{X}$, $\lambda$, $x_0$, and $s$ be defined as earlier (Eqn. \ref{eqn:yeo-johnson-invariant}).
Let $w_i \geq 0$ be the weight corresponding to each element of $\mathbf{X}$.
Let $\mu_w$ be the weighted mean of the Yeo-Johnson transformed sequence:
\begin{equation*}
\mu_w = \frac{\sum_{i=1}^n w_i \phi_{\text{YJ}}^{\lambda, x_0, s} (x_i)} {\sum_{i=1}^n w_i}
\end{equation*}

Let $\sigma^2_w$ be the weighted variance of the Yeo-Johnson transformed sequence:
\begin{equation*}
\sigma_w^2 = \frac{\sum_{i=1}^n w_i \left(\phi_{\text{YJ}}^{\lambda, x_0, s} (x_i) - \mu_w \right)^2}{\sum_{i=1}^n w_i}
\end{equation*}

Then, the weighted log-likelihood function of the location- and scale- invariant Yeo-Johnson power transformation is:

\begin{equation}
\label{eqn:yeo-johnson-weighted-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{rYJ}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi_{\text{YJ}}^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \left( \sum_{i=1}^n w_i \right) \log s + (\lambda - 1) \sum_{i=1}^n w_i \mathop{\mathrm{sgn}}(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}
\end{definition}

@@

\begin{itemize}
\item
  A step function, with \(\delta_1 \geq 0\) as threshold parameter:
  \begin{equation*}
  w_i =
  \begin{cases}
  1 & \text{if } \left| \dot{x}_i \right| \leq \delta_1\\
  0 & \text{if } \left| \dot{x}_i \right| > \delta_1
  \end{cases}
  \end{equation*}
\item
  A triangle function (or generalised Huber weight), with
  \(\delta_1 \geq 0\) and \(\delta_2 \geq \delta_1\) as threshold
  parameters: \begin{equation*}
  w_i =
  \begin{cases}
  1 & \text{if } \left| \dot{x}_i \right| < \delta_1\\
  1 - \frac{\left| \dot{x}_i \right| - \delta_1}{\delta_2 - \delta_1} & \text{if } \delta_1 \leq \left| \dot{x}_i \right| \leq \delta_2 \\
  0 & \text{if } \left| \dot{x}_i \right| > \delta_2
  \end{cases}
  \end{equation*}
\item
  A tapered cosine function (J. W. Tukey 1967), with \(\delta_1 \geq 0\)
  and \(\delta_2 \geq \delta_1\) as threshold parameters:
  \begin{equation*}
  w_i =
  \begin{cases}
  1 & \text{if } \left| \dot{x}_i \right| < \delta_1\\
  0.5 + 0.5 \cos\left(\pi \frac{\left| \dot{x}_i \right| - \delta_1}{\delta_2 - \delta_1} \right) & \text{if } \delta_1 \leq \left| \dot{x}_i \right| \leq \delta_2 \\
  0 & \text{if } \left| \dot{x}_i \right| > \delta_2
  \end{cases}
  \end{equation*}
\end{itemize}

@@

\caption{Weighting functions investigated in this study to make power transformations more robust against outliers. In this example, the step function was parameterised with $\delta_1 = 0.60$. The triangle and tapered cosine functions were both parameterised with $\delta_1 = 0.30$ and $\delta_2 = 0.90$.}\label{fig:weighting-functions}
\end{figure}

@@

\begin{itemize}
\item
  The weighting function uses empirical probabilities of the
  distribution of the original feature \(\mathbf{X}\). After sorting
  \(\mathbf{X}\) in ascending order, probabilities are determined as
  \(p_i = \frac{i - 1/3}{n + 1/3}\), with \(i = 1, 2, \ldots n\), with
  \(n\) the number of instances of feature \(\mathbf{X}\). Then
  \(\dot{x}_i = p^{*}_i=2 \left( p_i - 0.5\right)\), so that argument is
  zero-centered.
\item
  The weighting function uses the z-score of the transformed feature
  \(\phi^{\lambda, x_0, s} (\mathbf{X})\). After Raymaekers and
  Rousseeuw (2024),
  \(z_i = \frac{\phi^{\lambda, x_0, s}(x_i) - \mu_M}{\sigma_M}\). Here,
  \(\mu_M\) and \(\sigma_M\) are robust Huber M-estimates of location
  and scale of the transformed feature
  \(\phi^{\lambda, x_0, s} (\mathbf{X})\) (Huber 1981). Then
  \(\dot{x}_i = z_i\).
\item
  After sorting \(\mathbf{X}\) in ascending order, the weighting
  function uses the residual error between the z-score of the
  transformed feature \(\phi^{\lambda, x_0, s} (\mathbf{X})\) and the
  theoretical z-score from a standard normal distribution:
  \(r_i =\left| \left( \phi^{\lambda, x_0, s}(x_i) - \mu_M)\right) / \sigma_M - F^{-1}_{\mathcal{N}}(p_i) \right|\),
  with \(\mu_M\), \(\sigma_M\) and \(p_i\) as defined above. Then
  \(\dot{x}_i = r_i\).
\end{itemize}
